\documentclass{article}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amsthm}
\usepackage{stmaryrd}
\newcommand{\lens}{\leftrightarrow}
\newcommand{\slens}{\lens_s}
\newcommand{\R}{\mathbb R}
\newcommand{\cupdot}{\mathbin{\mathaccent\cdot\cup}}
\newcommand{\capdot}{\mathbin{\mathaccent\cdot\cap}}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{definition}{Definition}
\newtheorem{proposition}{Proposition}
\begin{document}
The goal is to take the theme ``bidirectional spreadsheet'' and run with it.
In particular we'd like to design something that would be helpful to
business people---who are familiar with normal unidirectional
spreadsheets---when they realize that a particular computation hasn't come
out the way they wanted it to, and are trying to figure out a way to guide
their business so that it comes out that way in the future.

(TODO: give a running example, so the drawbacks outlined in later sections
don't come as a surprise)

\section{Literal Interpretation}
\label{sec:naive}
One basic design idea is to simply describe the connections between cells
(called ``formulas'' in traditional spreadsheets) via a lens rather than a
one-way function. We can do this whenever:

\begin{enumerate}
    \item We know how to turn a particular math operator into a lens, that
        is, there's a ``good'' interpretation of inverting the operation.
    \item The formulas refer to each other in simple enough ways that we
        can use the monoidal structure of the category of lenses without
        relying on traced or cartesian structure (since we don't have
        those).
\end{enumerate}

In fact, when these conditions are satisfied, the lens laws tell us that
there is a \emph{unique} update to the spreadsheet induced by the lenses'
get and put functions.

To be specific, the second condition above can be formalized in the
following way. We identify a set $C$ of cell names, a set $T$ of types, and
a finite partial map $f \in C \to T$ giving the type of data stored in each
cell. We say a lens has spreadsheet-type $c_1 \times \cdots \times c_n
\slens c$ when $c, c_1, \ldots, c_n$ are pairwise distinct and the lens has
lens type $f(c_1) \times \cdots \times f(c_n) \lens f(c)$. Given a
collection of lenses and corresponding spreadsheet types, we can create a
directed graph whose nodes are $C$ and where there is an edge $(c,c')$ for
each lens whose type has the form $c_1 \times \cdots \times c' \times \cdots
\times c_n \slens c$. We say a collection of lenses has \emph{simple
structure} when there is at most one \emph{undirected} path between any pair
of nodes.

This is a fairly restrictive structural restriction, but note that it does
allow two (possibly surprising) setups: first, a given cell can have
multiple formulas associated with it, provided that they depend on disjoint
cell sets; second, multiple formulas can depend on a single cell, provided
the cells they output to are depended on by disjoint cell sets.

Suppose we now change the value in cell $c$. Pseudocode to update the rest
of the spreadsheet looks like this:

\newcommand\assign{\mathop{:=}}
\newcommand\modified{\mathit{modified}}
\newcommand\domain{\mathit{domain}}
\newcommand\codomain{\mathit{codomain}}
\begin{tabular}{l}
$\modified\assign\{c\}$ \\
iterate to fixpoint: \\
\qquad $k \assign $(nondeterministically) any lens \\
\qquad $f \assign $(nondeterministically) get or put \\
\qquad reject the choice when we would overwrite old work, i.e. when: \\
\qquad\qquad $k.f.\domain \cap \modified = \emptyset$ and $k.f.\codomain \cap \modified \ne \emptyset$ \\
\qquad run $k.f$ \\
\qquad $\modified \assign \modified \cup k.f.\codomain$
\end{tabular}

One should immediately ask whether this definition is really good, that is,
whether it terminates and is convergent. Showing these things relies
critically on the simple structure property, which gives us the crucial fact
that each cell changes its value at most once (in exactly the iteration
where it is added to the $\modified$ set).

(TODO: write down this argument carefully)

\newcommand{\tax}{\mathit{tax}}
\newcommand{\base}{\mathit{base}}
\newcommand{\total}{\mathit{total}}
The approach outlined above is simple to describe and implement, and it
reuses a lot of domain knowledge we have built up about lenses, but its
restrictions are strong enough that even fairly reasonable-looking
spreadsheets are excluded. For example, consider a spreadsheet with just
three cells named ``base'', ``tax'', and ``total''. We could imagine writing
a lens between base and tax that preserves the formula
$\tax=0.08*\base$, and similarly a lens between base, tax, and total
that preserves the formula $\total=\base+\tax$. However, the structure
restriction above is not satisfied in this spreadsheet: there are two paths
in the dependency graph from total to base (one that goes through tax and
one that doesn't).

Additionally, it isn't clear how to handle multi-update (that is, setting
the values of several cells simultaneously) or constrained update (that is,
fixing the values of some cells during the update process). The next two
sections discuss ways to address these shortcomings.

\section{Linear Algebra}
\label{sec:global-analysis}
Consider again the simple three-cell spreadsheet with two equations:
\begin{align*}
    \tax &= 0.08*\base \\
    \total &= \base+\tax
\end{align*}
We might observe that the $\total$ equation could be rewritten in terms of
just $\base$:
\begin{align*}
    \tax   &= 0.08*\base \\
    \total &= 1.08*\base
\end{align*}
The lenses implementing these two equations \emph{do} have the simple
structure property. One may wonder in exactly which circumstances such a
refactoring can be done -- and whether it could be done automatically. In
the following, we suggest one circumstance in which we can do quite well in
that direction.

Suppose for the moment that each lens' get is an affine function, and that
the dependency graph is acyclic. (Being acyclic is a significant relaxation
from the simple structure property, but being affine functions is a
significant strengthening of the ``any lens will do'' attitude from the
previous approach.) Call a cell a ``root cell'' if it has no out edges --
that is, none of the affine formulas in the spreadsheet output to root
cells. A simple argument shows that, given a cell in the spreadsheet, we
can write a affine formula which maps the values of root cells to the value
of the given cell. The argument goes by induction on the length of the
longest path from a root cell to the given one, and proceeds as in the
example above by substituting in affine formulas for each non-root variable
at each step. In fact, we can go a step farther: we can write affine
transformations from root cells to any set of cells. If we manage to give a
characterization of when these affine transformations can be
bidirectionalized, then we will have given an account of how to handle the
multi-update problem and relax the simple structure requirement in the parts
of the spreadsheet where only affine formulae are used.

Thus, we can now frame our problem in another way: what is the right way to
bidirectionalize an affine transformation? Accordingly, we will now step
away from spreadsheets and frame our discussion in linear algebra terms.

\newcommand{\argmin}[1]{\underset{#1}{\operatorname{argmin}}}
\newcommand{\lget}{\mathit{get}}
\newcommand{\lput}{\mathit{put}}
\newcommand{\x}{\mathbf x}
\newcommand{\y}{\mathbf y}
A function $\lget \in \R^m \to \R^n$ is affine exactly when there is a matrix
$M$ of dimension $n \times m$ and vector $\mathbf b \in \R^n$ such that $\lget(\x) =
M\x +\mathbf b$.  Affine functions are surjective (and hence bidirectionalizable)
just when $M$ has rank $n$, so we assume this. When $m=n$, $f$ is a
bijection. The put function in this case is particularly boring, because it
ignores the original source:
\[\lput(\x,\y) = M^{-1}\y\]
The more interesting case is when $m>n$, and where each $\y$ is therefore
the image of a nontrivial subspace of $\R^m$. There are many heuristics one
may choose to identify a particular point in this subspace; we choose the
specification:
\[\lput(\x,\y) = \argmin{\x',\lget(\x')=\y}||\x' - \x||\]

\newcommand{\transpose}{^\top}
\begin{theorem}
    There exists an $m \times n$ matrix $N$ such that
    \[\lput(\x,\y) = \x + N(\y - \lget(\x))\]
    satisfies the specification above.
\end{theorem}
\begin{proof}
    The intuition is that we wish to move as little as possible in
    source-space to match the move in target-space. This can be achieved by
    minimizing how far we move in the null space of $M$, since (exactly)
    these motions result in no motion in target-space.

    Take a basis $\{\x_1,\ldots,\x_{m-n}\}$ for the null space of $M$. Then
    we will take:
    \begin{align*}
        B &= \left[\begin{array}{c}
                \x_1\transpose \\
                \vdots \\
                \x_{m-n}\transpose
            \end{array}\right] \\
        N &= \left[\begin{array}{c}
                M \\
                B
            \end{array}\right]^{-1}
            \left[\begin{array}{c}
                I_n \\
                \mathbf 0_{m-n,n}
            \end{array}\right] \\
    \end{align*}
    We must now argue three things: that the square matrix in the definition
    of $N$ is invertible; that $N$ produces a put function that roundtrips;
    and that the put function produced by $N$ produces minimal changes. The
    definition above was crafted so that (assuming for the moment that $N$
    exists) we have $MN = I$, which is used to show the roundtrip property,
    and $BN = \mathbf 0$, which is used to show minimality.

    invertible: (TODO)

    roundtrip:
        \begin{align*}
            \lget(\lput(\x,\y))
                &= M(\x+N(\y-M\x-\mathbf b))+\mathbf b & \mbox{definition of }\lget,\lput \\
                &= MN\y + M\x + \mathbf b - MN(M\x + \mathbf b) & \mbox{rearranging terms} \\
                &= \y & MN=I
        \end{align*}

    minimal: (TODO)
\end{proof}

Pending research:
\begin{itemize}
    \item How do we solve constrained update ``optimally''?
    \item Can we generalize these observations? (What operations did we
        really need above? Can we be polymorphic over a theory that offers
        these operations? Are there other known instantiations of that
        theory?)
    \item how do the Shostak algorithm and Groebner bases fit in?
\end{itemize}

\section{Multiway Lenses}
% TODO: list all the ways to fix partiality
\label{sec:multiway}
The discussion in Section~\ref{sec:global-analysis} puts us in a slightly
uncomfortable situation: previously, our model could handle each lens having
exactly one output cell, but our analysis gave us the ability to treat
collections of cells as the output of a lens -- and indeed, to have a single
semantic object which could define lenses between the input cells and many
possible collections of output cells. Additionally, not all possible
collections of cells could act as outputs; in case the defining formulae for
the output cells were not linearly independent, the assumptions of the
analysis in terms of linear functions aren't satisfied. In this section, we
develop a more sophisticated semantic model which can capture these
additional behaviors. % TODO: motivate hiding


\newcommand{\Names}{\mathcal N}
\newcommand{\sto}{\mapsto}
\newcommand{\mlens}[1]{\mathcal M(#1)}
\newcommand{\mput}{\mathit{put}}
\newcommand{\danger}{\mathcal D}

Let $\Names$ be a set of cell names and 
$T$ a set of spreadsheet values like numbers or strings.

\begin{definition}
    A \emph{spreadsheet type} $N \subseteq \Names$ is a set of names.
\end{definition}

\begin{definition}
    An \emph{instance} $f$ of a spreadsheet type $N$ is a function $f : N \to T$.
\end{definition}

\begin{definition}
    Given a spreadsheet type $N$, a \emph{multiway lens} $\ell : \mlens N$
    is a quadruple $(N',\danger,\mput,K)$ where
    \begin{itemize}
        \item $N'$ is a spreadsheet type, called \emph{existential}, which is disjoint from $N$,
        \item $\danger : 2^{2^N}$ is a predicate on sets of names called the \emph{danger
            zone} of $\ell$,
        \item $\mput : 2^N \to (N \cup N' \to T) \to (N \cup N' \to T)$ is
            an update function which takes a set of names and an instance of
            the spreadsheet type $N \cup N'$ and produces another instance,
            and
        \item $K$ is an invariant expressed as a predicate on instances of
            $N \cup N'$.
    \end{itemize}
\end{definition}

\begin{definition}
    A multiway lens $\ell$ is \emph{well-formed} when it satisfies the
    following laws:
    \begin{itemize}
        \item $\danger$ is upwards-closed
        \item $\mput(i,f) \in K$ whenever $i \notin \danger$
        \item $\mput(i,f) = f$ whenever $f \in K$
        \item $\left.\mput(i,f)\right|_i = \left.f\right|_i$
    \end{itemize}
\end{definition}
When $i \notin \danger$ we say that $\mput(i,f)$ is \emph{well-scoped}.

\newcommand{\mdomain}[1]{\mathit{dom}(#1)}
\begin{definition}
    The \emph{domain} of a lens $\ell : \mlens N$, denoted $\mdomain\ell$, is
    $N$ itself, the set of cell names which represents the visible
    interface to the lens.
\end{definition}

Intuitively, we expect two lenses to be equivalent when they have the
same domain and their put functions behave identically on instances
restricted to that domain. However, we allow equivalent lenses to 
behave differently on their respective existential variables $C'$,
which need not be equal. 
\begin{definition}
    Lenses $k : \mlens N$ and $\ell : \mlens N$ are \emph{equivalent},
    written $k \equiv \ell$, when $k.\danger = \ell.\danger$ and there
    exists a relation $R$ between instances of $k.N'$ and $\ell.N'$ such
    that whenever (1) $f$ is an instance of $N$, (2) $g$ is an instance of $k.N'$
    and $h$ is an instance of $\ell.N'$ such that $g \mathrel R h$,
    and (3) $i\in N - k.\danger$, then there are some instances $f'$, $g'$ and
    $h'$ of $N$, $k.N'$ and $\ell.N'$ respectively such that
    \[k.\mput(i,f\cup g) = f' \cup g', \qquad
      \ell.\mput(i,f\cup h) = f' \cup h', \text{ and} \qquad
      g' \mathrel R h'. \]
\end{definition}

\begin{proposition}
    $\equiv$ is an equivalence relation.
\end{proposition}

The size of upward-closed elements of $2^{2^N}$ makes storing danger zones
in an implementation impractical. Instead, we use a compact representation
of danger zones such that the upward closure of the representation
spans the entire set.

\newcommand{\upclose}[2]{\llbracket #1 \rrbracket_{#2}}
\begin{definition}
    The \emph{upward closure} $\upclose VU$ of $V$ in $U$ is:
    \[S \in \upclose VU \iff \exists v \in V. v \subset S \subset U\]
\end{definition}

\begin{proposition}
    Upward closures are upward closed.
\end{proposition}


When a universe of names $U$ is understood, we will use the notation
\[E_{11}\cdots E_{1n_1},\cdots,E_{m1}\cdots E_{mn_m}\] 
as shorthand for
\[\upclose{\{\{E_{11},\cdots,E_{1n_1}\},\cdots,\{E_{m1},\cdots,E_{mn_m}\}\}}N.\]
For example, $AB,AC$ stands for $\{\{A,B\},\{A,C\},\{A,B,C\}\}$. In the definitions
below, we present $\danger$ using compact representations. For the lens
combinators in the following section, we construct danger zone 
representations from the representations of the sublenses. We define
union $\cupdot$ and intersection $\capdot$ operators over representations as 
\begin{align*}
    P \cupdot Q &= \{ t \mid t \in P \text{ or } t \in Q \} \\
    P \capdot Q &= \{ p \cup q \mid p \in P \text{ and } q \in Q \}
\end{align*}
so that $\upclose{R \cupdot S} U = \upclose{R} U \cup \upclose{S} U$ and
$\upclose{R \capdot S} U = \upclose{R} U \cap \upclose{S} U$.
 
\subsection{Combinators}

\newcommand{\cond}[1]{\left\{\begin{array}{ll}#1\end{array}\right.}

\newcommand{\focus}{\mathit{focus}}
If $k : \mlens{N}$ and $M$ is a spreadsheet type which is a superset of $N$,
then it is possible for the put functions of $k$ to act on an instance $f$
of $M$ even though $k.\mput$ is not well-defined on all cells in $M$. 
For the cells in $M \setminus N$, the extended put function is the identity.
\begin{definition} Let $M$ and $N$ be spreadsheet types with $N \subset M$. 
    Suppose $k : \mlens{N}$ is a lens, $i \in M$,
    and $f$ is an instance of $M$. Define $\focus(k,i,f)$ to be a spreadsheet
    instance of $M$ such that
    \begin{align*}
        \focus(k,i,f)(x) &= \cond{
            k.\mput(i \cap N,f|_N)(x) & x \in N \\
            f(x) & x \notin N
        }
    \end{align*}
\end{definition}
Notice that $\focus(k,i,f)$ is well-scoped if and only if $i \cap N \not\in k.\danger$.

Next, given two lenses $k$ and $\ell$ we will define a composition operation
which enforces the invariants of both $k$ and $\ell$.
\begin{definition}
    Suppose $k : \mlens{N_k}$ and $\ell : \mlens{N_\ell}$ such that 
    \[N_k \perp \ell.N' \perp k.N' \perp N_\ell,\]
    that is, $N_k \cap \ell.N' = \emptyset$, $\ell.N' \cap k.N' = \emptyset$,
    and $k.N' \cap N_\ell = \emptyset$. Let $T_k = N_k \cap k.N'$ and $T_\ell = N_\ell \cap \ell.N'$,
    and define $S = N_k \cap N_\ell$.
    Then the composition $k\mid\ell : \mlens{N_k \cup N_\ell}$ of $k$ and $\ell$ is
    \begin{align*}
        N' &= k.N' \cup \ell.N' \\
        \danger &= k.\danger \cup \ell.\danger \cup
            \{(d_k \cup d_\ell) \setminus S
            \mid d_k \in k.\danger, d_\ell \in \ell.\danger\} \\
        \mput(i) &= \cond{
            \focus(\ell,i\cup S) \circ \focus(k,i)
                & \mbox{ when well-scoped} \\
            \focus(k,i\cup S) \circ \focus(\ell,i)
                & \mbox{ otherwise}
            } \\
        K(f) &= k.K(f|_{T_k}) \land \ell.K(f|_{T_\ell})
    \end{align*}
\end{definition}

\begin{proposition}
    For all spreadsheet instances $f$ of $(N_k \cup N_\ell) \cup (k.N' \cup \ell.N')$, and
    for all $i \in N_k \cup N_\ell$ such that $i \not\in (k\mid\ell).\danger$,
    $(k\mid\ell).\mput(i,f)$ is well-scoped. In other words, at least one of
    $\focus(\ell,i\cup S) \circ \focus(k,i)$ or 
    $\focus(k,i\cup S) \circ \focus(\ell,i)$ is well-scoped.
\end{proposition}
\begin{proof}
    Suppose towards contradiction that both 
    $\focus(\ell,i\cup S) \circ \focus(k,i)(f)$
    and $\focus(k,i\cup S) \circ \focus(\ell,i)(f)$ are badly-scoped. 
    Then either
    \begin{itemize}
        \item $i \cap N_k \in k.\danger$;
        \item $i \cap N_\ell \in \ell.\danger$; or
        \item both $(i \cup S) \cap N_k \in k.\danger$ and $(i \cup S) \cap N_\ell \in \ell.\danger$ 
    \end{itemize}
    If $i \cap N_k \in k.\danger$, then $i \cap N_k \in (k\mid\ell).\danger$,
    but since danger zones are upwards closed, $i \in (k\mid\ell).\danger$ as well.
    Similarly, if $i \cap N_\ell \in \ell.\danger$ then $i \in (k\mid\ell).\danger$.
    In the third case, we know that
    \begin{align*}
        d_k &= (i \cup S) \cap N_k = (i \cap N_k) \cup S \in k.\danger \text{ and }\\
        d_\ell &= (i \cup S) \cap N_\ell = (i \cap N_\ell) \cup S \in \ell.\danger
    \end{align*}
    By the definition of $(k\mid\ell).\danger$, we know that
    $(d_k \cup d_\ell) \setminus S \in (k\mid\ell).\danger$.
    Consider that
    \begin{align*}
        d_k \cup d_\ell &= (i \cap N_k) \cup (i \cap N_\ell) \cup S \\
        (d_k \cup d_\ell) \setminus S &= ((i \cap N_k) \setminus S) \cup ((i \cap N_\ell) \setminus S) \\
        &= (i \cap (N_k \setminus N_\ell)) \cup (i \cap (N_\ell \setminus N_k)) \\
        &= i \cap ((N_k \setminus N_\ell) \cup (N_\ell \setminus N_k))
    \end{align*}
    But since this dangerous domain is a subset of $i$, $i$ is in the danger 
    zone of $(k\mid\ell).\danger$ as well.
\end{proof}

Conjecture: If $D_k$ and $D_\ell$ are the danger zone representations of 
    $k.\danger$ and $\ell.\danger$ respectively, where
    $k:\mlens {N_k}$ and $\ell:\mlens {N_\ell}$, then 
    \begin{align*}
        &\upclose {D_k \cup D_\ell \cup \{(d_k \cup d_l) \setminus (N_k \cup N_\ell) \mid d_k \in D_k \text{ and } d_\ell \in D_\ell\}} 
                {N_k \cup N_\ell} \\
        = &\upclose {D_k} {N_k} \cup \upclose {D_\ell} {N_\ell} \cup 
        \{(d_k \cup d_l) \setminus (N_k \cup N_\ell) \mid d_k \in \upclose{D_k} {N_k} 
                                           \text{ and } d_\ell \in \upclose{D_\ell} {N_\ell}\}.
    \end{align*}

We can extend composition to an arbitrary number of lenses. 
%TODO: discuss sequential composition vs multi-way composition
\begin{definition}
    The put function for a sequence of lenses $k_1,\ldots,k_n$ is
    \begin{align*}
        \mput_s(i) &= \focus\left(k_n,i \cup \bigcup_{m < n}\domain(k_m)\right)
            \circ \cdots \circ \focus\left(k_2, i \cup \domain(k_1)\right) \circ \focus(k_1,i)
    \end{align*}
\end{definition}
%TODO: define multi-way composition

In order to ensure that two lenses $k$ and $\ell$ have $k.D' \cap \ell.D' = \emptyset$,
we must be able to rename hidden cell names in a lens.
\begin{definition}
    Let $\ell : \mlens N$ where $V \in \ell.N'$ and $W \notin N \cup \ell.N'$. 
    Then the existential renaming $\ell\{W/V\} : \mlens N$ is
    \begin{align*}
        N' &= \{W\} \cup \ell.N' \setminus \{V\} \\
        \danger &= \ell.\danger \\
        \mput(i,f) &= \ell.\mput(i,f[V/w])[W/V] \\
        K(f) &= \ell.K(f[V/W])
    \end{align*}
\end{definition}
We can also define renaming on names in the lens's domain.
\begin{definition}
    Given $\ell : \mlens N$ and two names $V \in N$ and $W \notin N$, we can
    give the renaming $\ell[W/V]$:
    \begin{align*}
        N' &= \ell.N' \\
        \danger &= \{d[W/V] \mid d \in \ell.\danger\} \\
        \mput(i,f) &= \ell.\mput(i[V/W], f[V/W])[W/V] \\
        K(f) &= \ell.K(f[V/W])
    \end{align*}
\end{definition}
\begin{proposition}
    If $\ell\{W/V\}$ is well-defined, then $\ell\{W/V\} \equiv \ell$.
\end{proposition}

At times we will want to \emph{hide} or \emph{bind} a cell in a lens $k$, 
meaning that we want to move it from the external domain of $k$ to the
internal existential domain.
\begin{definition}
    Given $\ell : \mlens N$ and $V \in N$, we define $\nu V.\ell : \mlens{N
    \setminus \{V\}}$, pronounced ``hide $V$ in $\ell$,'' as:
    \begin{align*}
        N' &= \ell.N' \cup \{V\} \\
        \danger &= \{d \mid d \in \ell.\danger \land V \notin d\} \\
        \mput &= \ell.\mput \\
        K &= \ell.K
    \end{align*}
\end{definition}

\begin{proposition}
    Let $D$ be a danger zone representation of $\ell.\danger$ where
    $\ell : \mlens N$. Let $V \in N$. Then 
    \[ \upclose {\{d \mid d \in \ell.\danger \land V \notin d\}} {N\setminus\{V\}}
        = \{ d \mid d \in \upclose {\ell.\danger} N \land V \notin d\}. \]
\end{proposition}

\begin{proposition}
    Let $k : \mlens {N_k}$ and $\ell : \mlens {N_\ell}$ such that $N_k\perp\ell.N'\perp~k.N'\perp~N_\ell$.
    Let $V$ be name in $N_\ell$ such that $V \notin N_k$.
    Then $\nu V.(k \mid \ell) = k \mid (\nu V.\ell)$.
\end{proposition}
\begin{proof}
    Follows from the definition of hiding and composition.
    The domain of $\nu V.(k \mid \ell)$ is 
    $(N_k \cup N_\ell) \setminus \{V\} = N_k \cup (N_\ell \setminus \{V\})$ 
    since $V \notin N_k$, which is the domain of $k \mid (\nu V.\ell)$. 
    For the existential domains we have that
    \begin{align*}
        (\nu V.(k \mid \ell)).N' &= (k.N' \cup \ell.N') \cup \{V\} \text{ and }\\
        (k \mid (\nu V.\ell)).N' &= k.N' \cup (\ell.N' \cup \{V\})
    \end{align*}
    For the danger zone,
    \begin{align*}
        (\nu V.(k \mid \ell)).\danger &= \{d \in (k \mid \ell).\danger \mid V \notin d \} \\
        &= k.\danger \cup \{d \in \ell.\danger \mid V \notin d \} 
        \cup \{(d_k \cup d_\ell) \setminus S \mid V \notin (d_k \cup d_\ell)\setminus S \}
    \end{align*}
    Since $V \notin d_k \subseteq N_k$, and also that $V \notin S$, we know that
    $V \in (d_k \cup d_\ell) \setminus S$ if and only if $V \in d_\ell$. Therefore
    \begin{align*}
        (\nu V.(k \mid \ell)).\danger &= k.\danger \cup \{d \in \ell.\danger \mid V \notin d \}
        \cup \{(d_k \cup d_\ell) \setminus S \mid d_k \in k.\danger, d_\ell \in \ell.\danger, V \notin d_\ell \} \\
        &= k.\danger \cup (\nu V.\ell).\danger \cup \{(d_k \cup d_\ell) \setminus S \mid
            d_k \in k.\danger, d_\ell \in (\nu V.\ell).\danger \} \\
        &= (k \mid \nu V.\ell).\danger
    \end{align*}
    Since hiding has no effect on put functions or consistency relations, 
    this concludes the proof.
\end{proof}



\subsection{Primitive Lenses}


\newcommand{\id}{\mathit{id}}
\begin{definition}
    The identity lens $\id : \mlens{\{A,B\}}$ is defined as follows:
    \begin{align*}
        N' &= \{\} \\
        \danger &= AB \\
        \mput(\{A\},f) &= f[B \mapsto f(A)] \\
        \mput(\{B\},f) &= f[A \mapsto f(B)] \\
        \mput(\{\},f) &= f[A \mapsto f(B)] \\
        K(f) &= f(A) == f(B)
    \end{align*}
\end{definition} 

For a value $t \in T$, we define a constant lens
which enforces the invariant that a cell is always
equal to $t$.

\newcommand{\const}{\mathit{const}}
\begin{definition}
    If $t \in T$ then the constant lens
    $\const(t) : \mlens{\{A\}}$ is defined as:
    \begin{align*}
        N' &= \{\} \\
        \danger &= A \\
        \mput(i,f) &= \lambda v. t \\
        K(f) &= f(A) == t
    \end{align*}
\end{definition}

Conjecture: all these lenses and combinators are well-behaved.

Known false wish: $k \mid \ell \equiv \ell \mid k$ (possible fix:
nondeterminism)

Known false wish: $(k \mid \ell) \mid m \equiv k \mid (\ell \mid m)$
(possible fixes: track more detailed information about put method
interleavings, multiway composition, more restrictive danger zones)

Conjecture: when $k : A \lens B$ and $\ell : B \lens C$, we recover the
symmetric lens' $k;\ell : A \lens C$ via our $\nu B.(k \mid \ell)$.

(TODO: perhaps define a few other base lenses like affine transformations,
division/other operators, stringy stuff?)

Pending research:
\begin{itemize}
    \item more syntax and a big example
    \item other ideas we can take from process calculi? (e.g. name hiding,
        two different kinds of composition, etc.)
    \item types
    \item how do multicategories fit in?
\end{itemize}
\end{document}
